Quasi-universal acceptance of basic science in the US

### 2024-09-12 paper.Rmd
first version within word limit of 4k

appendix:
  - "appendix_exp1.Rmd"
  - "appendix_exp2.Rmd"
  - "appendix_exp3.Rmd"
  - "appendix_exp4.Rmd"

@@

Trust in science is related to many desirable outcomes, from acceptance of anthropogenic climate change (ref) or vaccination (ref) to following recommendations during COVID [@alganTrustScientistsTimes2021], which suggests that trust in science was the most important predictor of these behaviors).

@@

This relative lack of trust in science could mean different things. It could mean that some people reject science wholesale, or at least to a significant extent. Or it could mean that they maintain a high degree of trust in most of science, and only question specific results that happen to contradict specific other beliefs they have (e.g. creationism), or beliefs that are associated with disliked behaviors (e.g. vaccination, acting for climate change).

We answer this question by looking at how much Americans accept basic scientific knowledge. We pay special attention to two groups of participants: people who say they don’t trust science (much, or at all), and people who endorse science-related conspiracy theories (CTs), or have a CT mindset, since this has been linked with a low degree of trust in science, and that many CTs imply a mistrust of at least some science (vaccination, hydroxychloroquine, etc.).

We briefly review work on trust in science, and science knowledge, before turning to work on CT and trust in science.

## Trust in science

Across the globe, most people say they trust science at least to some extent. In 2018, the Wellcome Global Monitor (WGM) surveyed of over 140'000 people in over 140 countries on trust in science [@wellcomeglobalmonitorWellcomeGlobalMonitor2018]. In 2020, during the first year of the Covid pandemic, a follow up survey was made in 113 countries, involving 119'000 participants [@wellcomeglobalmonitorWellcomeGlobalMonitor2020]. Across all countries, in 2018, 32% of participants said they trust science "a lot" (41% in 2020), 45% trust "some" science (39% in 2020), and only 13% percent trust science "not much/ not at all" (13% in 2020), while 10% indicated "don't know" (7% in 2020)[^1].

[^1]: <https://wellcome.org/reports/wellcome-global-monitor-covid-19/2020#gid=6acd&pid=0>\>

Trust in science also appears to be relatively stable in the US [@funkPublicConfidenceScientists2020]. However, there are some potentially worrying trends: Recently, trust in US has dropped [@brianAmericansTrustScientists2023], and some evidence suggests increasing polarization of trust in science along political lines [@gauchatPoliticizationSciencePublic2012]. For example, the partisan divide on the question of whether climate change is a major threat to society has been growing in the last decade[^2] [@kennedyWhatDataSays2023]. A recent study suggests that by some measure, only about half of the US population believes in anthropogenic climate change, and that individuals do mostly not seem to change their mind on on the issue [@mottaChangingMindsChanging2021].

[^2]: <https://www.pewresearch.org/short-reads/2023/08/09/what-the-data-says-about-americans-views-of-climate-change/>

Trust in science is typically measured in three ways: either asking explicitly about general trust in science or scientists, by asking about science attitudes (e.g. support for public funding of science), or asking questions about specific, typically contentious, science topics (e.g. vaccines, climate change, GMOs). It is unclear to which extent these measures capture trust in basic science. To assess this, we can look at the literature on science knowledge.

[introduce SK as: that gives us a lower bound of how much people trust basic science]

## Science knowledge

Much of the research on public understanding of science measures to which extent people have basic science knowledge (e.g. "Does the Earth go around the Sun, or does the Sun go around the Earth?"). For us, these surveys provide a lower bound to how much people trust basic science: people who give the right answer (and not by chance) supposedly trust science on that question. Unfortunately, that lower bound has often been indeed pretty low: researchers have pointed out the deficit in basic science knowledge [@millerPublicUnderstandingAttitudes2004; @millerMeasurementCivicScientific1998a]. For example, only 55% in the US (in 2014) and only 46% in the EU (in 2005) knew that antibiotics kill only bacteria and not viruses [@committeeonscienceliteracyandpublicperceptionofscienceScienceLiteracyConcepts2016]. A question that remains is whether people do not provide the right answers because they are not familiar with the scientific consensus, or because they explicitly reject that consensus.

[To do: Make a descriptive overview of science knowledge across time/countries.]

## CT and TiS

Recently, the relationship of one group to trust in science (TiS) has been particularly scrutinized: conspiracy theorists (CTists), either defined as people who endorse specific conspiracy theories, or as people who have a conspiracy mindset [@rutjensConspiracyBeliefsScience2022; @vranicDidMyOwn2022]. CTists reject specific science knowledge (related to their CT), and they tend to trust science less on the whole. What remains unclear is to which extent CTists (dis)trust basic science.

@@

In a series of four pre-registered online studies (total n = 782), we asked US participants questions about well-established scientific facts. For each question, we asked participants what they thought the correct answer was (testing their knowledge of science), we informed them of the scientifically accepted correct answer, and asked them whether they accepted it (measuring their trust in basic science). We also measured participants’ trust in science using standard measures, as well as their beliefs in various conspiracy theories and their tendency to engage in conspiratorial thinking. The four studies, including materials, hypotheses, and analyses, were pre-registered and all materials and data are accessible via the Open Science Framework (OSF)). The differences between the four studies are summarized presently, and the methods are detailed below.

@@

In Study 1, we used questions drawn from questionnaires of scientific knowledge (e.g. "Are electrons smaller, larger, or the same size as atoms? [Smaller; Same size; Larger]"), supplemented by a ‘trick’ question ("Where do trees mainly draw the materials with which they create their mass? [Earth; Water; Air]"; correct answer: Air). In Studies 2 and 3, this last question was removed. The scientific facts used in Studies 1 to 3 represent long-established and basic knowledge. In Study 4 we used more recent much less basic scientific discoveries (e.g. "How many more glial cells are there in the brain in comparison with neurons, as established in 2016? [The same amount; Twice as many; Tenth as many]"; correct answer: The same amount).

@@

In Study 1, we simply told participants that they would be provided with the scientifically consensual answer. However, for participants to accept this answer, they must not only trust science, but also trust us (that we are presenting them with the actual scientifically consensual answer). To remove this issue, in Studies 2 to 3, we presented participants with a short explanation of the correct answer, as well as links to three sources per answer (e.g. Wikipedia, National Geographic or NASA). In Study 4, as the topics were more complex, we did not provide an explanation, but we provided two sources per answer.

@@

# Exploratory: Differences by reason for trust
trust_by_reason_model <- run_regression(exp3_wide, dependent_var = "wgm_sciencegeneral", independent_var = "reason_agreement")

acceptance_by_reason_model <-run_regression(exp3_wide, dependent_var = "avg_acceptance", independent_var = "reason_agreement")

# extract descriptives for inline reporting
exp3_descriptives <- list(
  # Demographics
  n_subj = n_distinct(exp3_wide$id),
  n_subject_by_outcome = exp3_wide %>% 
    summarize(across(c("wgm_sciencegeneral", "reason_agreement", "BCTI_avg"), 
                     ~sum(!is.na(.x)
                     )
    )
    ),
  gender = exp3_wide %>% group_by(gender) %>% summarize(n = n_distinct(id)) %>% split(.$gender),
  age = exp3_wide %>% summarize(across(age,list(mean = mean, median = median, sd = sd), 
                                       .names = "{.fn}")) %>% rounded_numbers(),
  # RQ1 & RQ2
  means = exp3_wide %>% 
    summarize(across(c(avg_knowledge, avg_acceptance, avg_acceptance_initial), 
                     list(mean = mean, sd = sd), 
                     .names = "{.col}_{.fn}")) %>%
    rename_with(~sub("^avg_", "", .), everything()) %>% 
    mutate_if(is.numeric, round, digits = 2) %>% 
  mutate(across(ends_with("_mean"), ~paste0(.*100, " %"))) ,
  means_numeric = exp3_wide %>% 
    summarize(across(c(avg_knowledge, avg_acceptance, avg_acceptance_initial), 
                     list(mean = mean, sd = sd), 
                     .names = "{.col}_{.fn}")) %>%
    rename_with(~sub("^avg_", "", .), everything()),
  # H1
  # a
  cor_trust_knowledge = cor_trust_knowledge,
  # b (conditional on false responses)
  false_answers_cor_trust_acceptance = false_answers_cor_trust_acceptance,
  # b (all cases)
  cor_trust_acceptance = cor_trust_acceptance,
  # H2
  # a
  cor_conspiracy_knowledge = cor_conspiracy_knowledge,
  # b (conditional on false responses)
  false_answers_cor_conspiracy_acceptance = false_answers_cor_conspiracy_acceptance,
  # b(all cases)
  cor_conspiracy_acceptance = cor_conspiracy_acceptance, 
  # Exploratory
  # acceptance and knowledge
  conditional_acceptance = exp3_long %>% 
    pivot_longer(c(acceptance, acceptance_initial), 
                 names_to = "measure", 
                 values_to = "acceptance") %>% 
    group_by(knowledge, measure, acceptance) %>% 
    count() %>% 
    group_by(knowledge, measure) %>% 
    mutate (share = n/sum(n), 
            # rename knowledge values
            knowledge = ifelse(knowledge == FALSE, "false", "true")) %>% 
    mutate_if(is.numeric, round, digits = 3) %>% 
    mutate (share = paste0(share*100, " %")) %>% 
    super_split(knowledge,  measure, acceptance),
  # Justifications
  justifications_n = nrow(exp3_justifications), 
  justifications_n_participants = n_distinct(exp3_justifications$id),
  justification_by_category = exp3_justifications %>%
    group_by(category)%>%
    summarize(n = n(), 
              n_subjects = n_distinct(id)) %>%
    mutate(share = n/sum(n)) %>%
    mutate_if(is.numeric, round, digits=3) %>% 
    mutate(share = paste0(share*100, "%")) %>% 
    split(.$category),
  # Consensus acceptance
  acceptance_n = exp3_wide %>% 
    filter(!is.na(reason_agreement)) %>% 
    nrow(.),
  acceptance_by_reason = exp3_wide %>% 
    group_by(reason_agreement) %>% 
    summarize(n = n()) %>%
    drop_na(reason_agreement) %>% 
    mutate(share = n/sum(n)) %>%
    mutate_if(is.numeric, round, digits=3) %>% 
    mutate(share = paste0(share*100, "%")) %>% 
    split(.$reason_agreement),
  reason_followup_n = exp3_wide %>% 
    filter(!is.na(reason_followup)) %>% 
    nrow(.),
  # Exploratory: Differences by reason for trust
  reason_means = exp3_wide %>% 
    group_by(reason_agreement) %>% 
    summarize(across(c(wgm_sciencegeneral, avg_acceptance), 
                     list(mean = mean, sd = sd), 
                     .names = "{.col}_{.fn}")) %>%
    rename_with(~sub("^avg_", "", .), everything()) %>% 
    mutate_if(is.numeric, round, digits = 2) %>% 
    mutate(across(acceptance_mean, ~paste0(.*100, " %"))) %>% 
    split(.$reason_agreement), 
  reason_trust_model = trust_by_reason_model %>% split(.$term), 
  reason_acceptance_model = acceptance_by_reason_model %>% split(.$term)
)
```

@@

# Exploratory: Differences by reason for trust
trust_by_reason_model <- run_regression(exp4_wide, dependent_var = "wgm_sciencegeneral", independent_var = "reason_agreement")

acceptance_by_reason_model <- run_regression(exp4_wide, dependent_var = "avg_acceptance", independent_var = "reason_agreement")

clicks_by_reason_model <- run_regression(exp4_wide, dependent_var = "sum_link_clicks", independent_var = "reason_agreement")

@@

    summarize(across(c(wgm_sciencegeneral, avg_acceptance, sum_link_clicks), 

@@

  reason_clicks_model = clicks_by_reason_model %>% split(.$term)

@@

```{r combined-data}
# combine data of all three studies
combined_data <- bind_rows(exp1_long %>% 
                             mutate(study = "1"), 
                           exp2_long %>% 
                             mutate(study = "2"), 
                           exp3_long %>% 
                             mutate(study = "3"),
                           exp4_long %>% 
                             mutate(study = "4")
                           )

@@

# extract descriptives for inline reporting
combined_descriptives <- list(
  # acceptance
  mean_acceptance = combined_data %>% 
  group_by(acceptance) %>% 
  count() %>% 
    ungroup() %>% 
    mutate(share = n/sum(n)) %>% 
    mutate_if(is.numeric, round, digits = 3) %>% 
    mutate(share = paste0(share*100, " %")) %>% 
    split(.$acceptance), 
  # acceptance by trust
  acceptance_by_trust = combined_data %>% 
    group_by(acceptance, wgm_sciencegeneral) %>% 
    count() %>% 
    group_by(wgm_sciencegeneral) %>% 
    mutate(share = n/sum(n)) %>% 
    mutate_if(is.numeric, round, digits = 3) %>% 
    mutate(share = paste0(share*100, " %")) %>% 
    super_split(wgm_sciencegeneral, acceptance), 
  participants_by_trust = combined_data %>% 
  group_by(wgm_sciencegeneral) %>% 
  summarize(n_participants = n_distinct(id)) %>% 
    mutate(share = n_participants/sum(n_participants)) %>% 
    mutate_if(is.numeric, round, digits = 3) %>% 
    mutate(share = paste0(share*100, " %")) %>% 
    split(.$wgm_sciencegeneral),
  # acceptance by conspiracy thinking
  acceptance_by_conspiracy_thinking = combined_data %>% 
    group_by(acceptance, CMQ_avg_binned) %>% 
    count() %>% 
    group_by(CMQ_avg_binned) %>% 
    mutate(share = n/sum(n)) %>% 
    mutate_if(is.numeric, round, digits = 3) %>% 
    mutate(share = paste0(share*100, " %")) %>% 
    super_split(CMQ_avg_binned, acceptance),
  participants_by_conspiracy_thinking  = combined_data %>% 
    group_by(CMQ_avg_binned) %>% 
    summarize(n_participants = n_distinct(id)) %>% 
    mutate(share = n_participants/sum(n_participants)) %>% 
    mutate_if(is.numeric, round, digits = 3) %>% 
    mutate(share = paste0(share*100, " %")) %>% 
    split(.$CMQ_avg_binned),
  # acceptance by conspiracy belief
  acceptance_by_conspiracy_belief = combined_data %>% 
    # bring to long format
    pivot_longer(cols = BCTI_apollo:BCTI_evolution, 
                 names_to = "BCTI_item", values_to = "score") %>% 
    mutate(BCTI_item = sub("^BCTI_", "", BCTI_item)) %>% 
    group_by(acceptance, BCTI_item, score) %>% 
    count() %>% 
    group_by(BCTI_item, score) %>% 
    mutate(share = n/sum(n)) %>% 
    mutate_if(is.numeric, round, digits = 3) %>% 
    mutate(share = paste0(share*100, " %")) %>% 
    super_split(acceptance, BCTI_item, score),
  participants_by_conspiracy_belief  = combined_data %>% 
    # bring to long format
    pivot_longer(cols = BCTI_apollo:BCTI_evolution, 
                 names_to = "BCTI_item", values_to = "score") %>% 
    mutate(BCTI_item = sub("^BCTI_", "", BCTI_item)) %>% 
    group_by(BCTI_item, score) %>% 
    summarize(n_participants = n_distinct(id)) %>% 
    mutate(share = n_participants/sum(n_participants)) %>% 
    mutate_if(is.numeric, round, digits = 3) %>% 
    mutate(share = paste0(share*100, " %")) %>% 
    super_split(BCTI_item, score),
  # knowledge
  mean_knowledge = combined_data %>% 
    group_by(knowledge) %>% 
    count() %>% 
    ungroup() %>% 
    mutate (share = n/sum(n), 
            # rename knowledge values
            knowledge = ifelse(knowledge == FALSE, "false", "true")) %>% 
    mutate_if(is.numeric, round, digits = 3) %>% 
    mutate(share = paste0(share*100, " %")) %>% 
    split(.$knowledge)
)
```

@@

For study 2, we restricted our main hypotheses about acceptance to cases in which participants initially provided a wrong answer. The rationale was to see if trust in science or conspiracy belief were associated with a change of mind towards the consensus. We provide results on these conditional correlations--for Study 2 and for all other studies--in the ESM[^3]. However, for the analysis presented here, we proceeded as preregistered for all other studies, by reporting simple, unconditional correlations between acceptance and trust in science, or, respectively, conspiracy belief.

[^3]: Only in study 4 do we find evidence that changing one's mind towards the scientific consensus is associated with (more) trust in science (study 1: r = `r exp1_descriptives$false_answers_cor_trust_acceptance$estimate`, p `r exp1_descriptives$false_answers_cor_trust_acceptance$p.value`; study 2: r = `r exp2_descriptives$false_answers_cor_trust_acceptance$estimate`, p `r exp2_descriptives$false_answers_cor_trust_acceptance$p.value`; study 3: r = `r exp3_descriptives$false_answers_cor_trust_acceptance$estimate`, p `r exp3_descriptives$false_answers_cor_trust_acceptance$p.value`; study 4: r = `r exp4_descriptives$false_answers_cor_trust_acceptance$estimate`, p `r exp4_descriptives$false_answers_cor_trust_acceptance$p.value`) and only in study 2 evidence that it is associated with (less) conspiracy beliefs and (less) conspiracy thinking (study 1: r = `r exp1_descriptives$false_answers_cor_conspiracy_acceptance$estimate`, p `r exp1_descriptives$false_answers_cor_conspiracy_acceptance$p.value`; study 2: r = `r exp2_descriptives$false_answers_cor_conspiracy_acceptance$estimate`, p `r exp2_descriptives$false_answers_cor_conspiracy_acceptance$p.value`; study 3: r = `r exp3_descriptives$false_answers_cor_conspiracy_acceptance$estimate`, p `r exp3_descriptives$false_answers_cor_conspiracy_acceptance$p.value`; study 4: r = `r exp4_descriptives$false_answers_cor_conspiracy_acceptance$estimate`, p `r exp4_descriptives$false_answers_cor_conspiracy_acceptance$p.value`).

@@

After providing their consent to participate in the study, participants were given an attention check "While watching the television, have you ever had a fatal heart attack?" [1-6; 1 = Never, 6 = Often]. All participants who did not answer "1 = Never" were excluded. Participants then read the following instructions:"We will ask you 10 questions about science. After each question, we will provide you with the scientifically consensual answer and ask whether you accept it." Next, participants answered a set of 10 basic science questions (in Study 1 they were randomly selected from a pool of 11 questions, as one ‘trick’ question had been added) in random order. After each question, participants were presented with an answer reflecting the scientific consensus, and asked whether they accepted it. In Studies 2 and 3, participants additionally saw a short explanation, partly based on explanations generated by ChatGPT, and three links to authoritative sources supporting the answer. In Study 4, we provided only two links and no explanation. Participants then answered questions on conspiracy thinking, conspiracy beliefs, and trust in science.

@@

After removing failed attention checks, we were left with a total sample size of `r exp1_descriptives$n_subj + exp2_descriptives$n_subj + exp3_descriptives$n_subj + exp4_descriptives$n_subj` (`r exp1_descriptives$n_subj` in Study 1, 6 failed attention checks; `r exp2_descriptives$n_subj` in Study 2, 11 failed attention checks; `r exp3_descriptives$n_subj` in Study 3, no failed attention checks; `r exp4_descriptives$n_subj` in Study 4, 2 failed attention checks) participants in the US via prolific. Details and demographics can be found in the online supplemental material. While samples for Studies 1 and 2 were convenience samples, Studies 3 and 4 were conducted on a sample holding vaccine-skeptic beliefs. Prolific Academic allows to select participants based on their answers to a range of questions. We picked three of these questions and only recruited participants who met our criteria for each of them:

1.  "Please describe your attitudes towards the COVID-19 (Coronavirus) vaccines: [For (I feel positively about the vaccines); Against (I feel negatively about the vaccines); Neutral (I don't have strong opinions either way); Prefer not to say"]. We selected particpants who answered "Against".
2.  "Have you received a coronavirus (COVID-19) vaccination? [Yes (at least one dose); No; Prefer not to answer]". We select only people who answered "No".
3.  "On a scale from 1-7, please rate to what extent you agree with the following statement: I believe that scheduled immunizations are safe for children. [1 (totally disagree); 2 (disagree); 3 (somewhat disagree); 4 (neither agree nor disagree); 5 (somewhat agree); 6 (agree); 7 (totally agree); rather not say]". We select only people who answered '1', '2', or '3'.

@@

We selected 10 science/health related conspiracy theories from the Belief in Conspiracy Theory Inventory (BCTI) [@pennycookOverconfidentlyConspiratorialConspiracy2022] (Table \@ref(tab:conspiracy)). Participants were asked: "Below is a list of events for which the official version has been disputed. For each event, we would like you to indicate to what extent you believe the cover-up version of events is true or false. [1-9; labels: 1 - completly false, 5 - unsure, 9 - completely true]".

@@

>>>>

appendix:
  - "appendix_exp1.Rmd"
  - "appendix_exp2.Rmd"
  - "appendix_exp3.Rmd"
  - "appendix_exp4.Rmd"
  - "appendix_acceptance_knowledge.Rmd"

@@

Trust in science is related to many desirable outcomes, from acceptance of anthropogenic climate change [@colognaRoleTrustClimate2020] or vaccination [@sturgisTrustScienceSocial2021; @lindholtPublicAcceptanceCOVID192021] to following recommendations during COVID [@alganTrustScientistsTimes2021], which suggests that trust in science was the most important predictor of these behaviors).

@@

@@

In a series of four pre-registered online studies (total n = 782), we asked US participants questions about well-established scientific facts. For each question, we asked participants what they thought the correct answer was (testing their knowledge of science), we informed them of the scientifically accepted answer, and asked them whether they accepted it (measuring their trust in basic science). We also measured participants’ trust in science using standard measures, as well as their beliefs in various conspiracy theories and their tendency to engage in conspiratorial thinking. The four studies, including materials, hypotheses, and analyses, were pre-registered and all materials and data are accessible via the Open Science Framework (OSF). The differences between the four studies are summarized presently, and the methods are detailed below.

@@

In Study 1, we used questions drawn from questionnaires of scientific knowledge (e.g. "Are electrons smaller, larger, or the same size as atoms? [Smaller; Same size; Larger]"), supplemented by a ‘trick’ question ("Where do trees mainly draw the materials with which they create their mass? [Earth; Water; Air]"; correct answer: Air). In Studies 2 and 3, this last question was removed. The scientific facts used in Studies 1 to 3 represent long-established and basic knowledge. In Study 4 we used more recent much less basic scientific discoveries (e.g. "What is the electric charge of the Higgs Boson, as established in 2012? [1.602176634 × 10-19; 0; 3.2×10-19C]"; correct answer: 0, i.e. electrically neutral).

@@

In Study 1, we simply told participants that they would be provided with the scientifically consensual answer. However, for participants to accept this answer, they must not only trust science, but also trust that we are presenting them with the actual scientifically consensual answer. To remove this issue, in Studies 2 to 3, we presented participants with a short explanation of the correct answer, as well as links to three sources per answer (e.g. Wikipedia, National Geographic or NASA). In Study 4, as the topics were more complex, we did not provide an explanation, but still provided two sources per answer.

@@

# Exploratory: Differences by reason

# trust
trust_by_reason_model <- run_regression(exp3_wide, dependent_var = "wgm_sciencegeneral", independent_var = "reason_agreement")

# acceptance
acceptance_by_reason_model <-run_regression(exp3_wide, dependent_var = "avg_acceptance", independent_var = "reason_agreement")

# BCTI (conspiracy beliefs)
BCTI_by_reason_model <- run_regression(exp3_wide, dependent_var = "BCTI_avg", independent_var = "reason_agreement")

# survey time
time_by_reason_model <- run_regression(exp3_wide %>% filter(duration_mins < 30), dependent_var = "duration_mins", independent_var = "reason_agreement")
exp3_descriptives <- list(
  n_subj = n_distinct(exp3_wide$id),
  n_subject_by_outcome = exp3_wide %>% 
  gender = exp3_wide %>% group_by(gender) %>% summarize(n = n_distinct(id)) %>% split(.$gender),
  age = exp3_wide %>% summarize(across(age,list(mean = mean, median = median, sd = sd), 
  means = exp3_wide %>% 
  means_numeric = exp3_wide %>% 
  conditional_acceptance = exp3_long %>% 
  justifications_n = nrow(exp3_justifications), 
  justifications_n_participants = n_distinct(exp3_justifications$id),
  justification_by_category = exp3_justifications %>%
  acceptance_n = exp3_wide %>% 
  acceptance_by_reason = exp3_wide %>% 
  reason_followup_n = exp3_wide %>% 
  reason_means = exp3_wide %>% 
    summarize(across(c(wgm_sciencegeneral, avg_acceptance, BCTI_avg), 
                     list(mean = ~mean(.x, na.rm = TRUE), 
                          sd = ~sd(.x, na.rm = TRUE)), 
  reason_BCTI_model = BCTI_by_reason_model %>%  split(.$term),
  reason_time_model = time_by_reason_model %>%  split(.$term), 
  n_participants_over30mins = exp3_wide %>% 
    filter(duration_mins >= 30) %>% 
    nrow(.)

@@

# Exploratory: Differences by reason

# trust
trust_by_reason_model <- run_regression(exp4_wide, dependent_var = "wgm_sciencegeneral", independent_var = "reason_agreement")

# acceptance
acceptance_by_reason_model <-run_regression(exp4_wide, dependent_var = "avg_acceptance", independent_var = "reason_agreement")

# BCTI (conspiracy beliefs)
BCTI_by_reason_model <- run_regression(exp4_wide, dependent_var = "BCTI_avg", independent_var = "reason_agreement")

# survey time
time_by_reason_model <- run_regression(exp4_wide %>% filter(duration_mins < 30), dependent_var = "duration_mins", independent_var = "reason_agreement")

# clicks
clicks_by_reason_model <- run_regression(exp4_wide, dependent_var = "sum_link_clicks", independent_var = "reason_agreement")

@@

    summarize(across(c(wgm_sciencegeneral, avg_acceptance, sum_link_clicks, BCTI_avg), 

@@

  reason_clicks_model = clicks_by_reason_model %>% split(.$term),
  reason_BCTI_model = BCTI_by_reason_model %>%  split(.$term),
  reason_time_model = time_by_reason_model %>%  split(.$term), 
  n_participants_over30mins = exp4_wide %>% 
    filter(duration_mins >= 30) %>% 
    nrow(.)

@@

```{r combined-data}
# combine data of all three studies
combined_data <- bind_rows(exp1_long %>% 
                             mutate(study = "1"), 
                           exp2_long %>% 
                             mutate(study = "2"), 
                           exp3_long %>% 
                             mutate(study = "3"),
                           exp4_long %>% 
                             mutate(study = "4")
                           ) %>% 
  # make a unique id variable
  mutate(id = paste0("study", study, "_", id))

@@

combined_descriptives <- list(
  # acceptance
  mean_acceptance = combined_data %>% 
  group_by(acceptance) %>% 
    ungroup() %>% 
    mutate(share = paste0(share*100, " %")) %>% 
    split(.$acceptance), 
  # acceptance by trust
  acceptance_by_trust = combined_data %>% 
    group_by(acceptance, wgm_sciencegeneral) %>% 
    group_by(wgm_sciencegeneral) %>% 
    mutate(share = paste0(share*100, " %")) %>% 
    super_split(wgm_sciencegeneral, acceptance), 
  participants_by_trust = combined_data %>% 
  group_by(wgm_sciencegeneral) %>% 
  summarize(n_participants = n_distinct(id)) %>% 
    mutate(share = n_participants/sum(n_participants)) %>% 
    mutate(share = paste0(share*100, " %")) %>% 
    split(.$wgm_sciencegeneral),
  # acceptance by conspiracy thinking
  acceptance_by_conspiracy_thinking = combined_data %>% 
    group_by(acceptance, CMQ_avg_binned) %>% 
    group_by(CMQ_avg_binned) %>% 
    mutate(share = paste0(share*100, " %")) %>% 
    super_split(CMQ_avg_binned, acceptance),
  participants_by_conspiracy_thinking  = combined_data %>% 
    group_by(CMQ_avg_binned) %>% 
    summarize(n_participants = n_distinct(id)) %>% 
    mutate(share = n_participants/sum(n_participants)) %>% 
    mutate(share = paste0(share*100, " %")) %>% 
    split(.$CMQ_avg_binned),
  # acceptance by conspiracy belief
  acceptance_by_conspiracy_belief = combined_data %>% 
    # bring to long format
    pivot_longer(cols = BCTI_apollo:BCTI_evolution, 
                 names_to = "BCTI_item", values_to = "score") %>% 
    mutate(BCTI_item = sub("^BCTI_", "", BCTI_item)) %>% 
    group_by(acceptance, BCTI_item, score) %>% 
    group_by(BCTI_item, score) %>% 
    mutate(share = paste0(share*100, " %")) %>% 
    super_split(acceptance, BCTI_item, score),
  participants_by_conspiracy_belief  = combined_data %>% 
    # bring to long format
    pivot_longer(cols = BCTI_apollo:BCTI_evolution, 
                 names_to = "BCTI_item", values_to = "score") %>% 
    mutate(BCTI_item = sub("^BCTI_", "", BCTI_item)) %>% 
    group_by(BCTI_item, score) %>% 
    summarize(n_participants = n_distinct(id)) %>% 
    mutate(share = n_participants/sum(n_participants)) %>% 
    mutate(share = paste0(share*100, " %")) %>% 
    super_split(BCTI_item, score),
  # knowledge
  mean_knowledge = combined_data %>% 
    group_by(knowledge) %>% 
    ungroup() %>% 
    mutate(share = paste0(share*100, " %")) %>% 
    split(.$knowledge),
  # flat earthers with no trust at all in science
  # number of participants
  flat_earth_no_trust_nparticipants = combined_data %>% 
    filter(BCTI_flatearth == 9 & wgm_sciencegeneral == 1) %>%
    summarize(n_participants = n_distinct(id)) %>% 
    pull(),
  # acceptance
  flat_earth_no_trust_acceptance = combined_data %>% 
    filter(BCTI_flatearth == 9 & wgm_sciencegeneral == 1) %>%
    group_by(acceptance) %>% 
    summarise(n = n()) %>% 
    mutate(share = paste0(share*100, " %")) %>% 
    split(.$acceptance), 
  # Lowest decile acceptance
  lowest_decile_acceptance = combined_data %>% 
    group_by(id) %>% 
    summarise(acceptance_rate = sum(acceptance == "Yes")/n(), 
              n = n()) %>% 
    arrange(acceptance_rate) %>% 
    slice(1:(n() * 0.1)) %>%  # Get the lowest 10% (decile)
    summarize(mean_acceptance_rate = mean(acceptance_rate, na.rm = TRUE), 
              n_smaller_or_equalto_0.5 = sum(acceptance_rate <=0.5)) %>% 
    mutate(mean_acceptance_rate = paste0(mean_acceptance_rate*100, " %")), 
  # Lowest decile knowledge
  lowest_decile_knowledge = combined_data %>% 
    group_by(id) %>% 
    summarise(acceptance_rate = sum(acceptance == "Yes")/n(), 
              knowledge_rate = mean(knowledge, na.rm=TRUE),
              n = n()) %>% 
    arrange(knowledge_rate) %>% 
    slice(1:(n() * 0.1)) %>%  # Get the lowest 10% (decile)
    summarize(mean_acceptance_rate = mean(acceptance_rate, na.rm = TRUE), 
              mean_knowledge_rate = mean(knowledge_rate, na.rm = TRUE)) %>% 
    mutate(mean_acceptance_rate = paste0(mean_acceptance_rate*100, " %"), 
           mean_knowledge_rate = paste0(mean_knowledge_rate*100, " %"))

@@

For Study 2, we restricted our main hypotheses about acceptance to cases in which participants initially provided a wrong answer. However, this meant the more participants had initially provided correct answers, the fewer opportunities they had for accepting correct answers. We provide results on these conditional correlations--for Study 2 and for all other studies--in the ESM[^1]. However, for the analysis presented here, we proceeded as preregistered for all other studies, by reporting unconditional correlations between acceptance and trust in science, or, respectively, conspiracy belief.

[^1]: Only in Study 4 do we find evidence that changing one's mind towards the scientific consensus is associated with (more) trust in science (Studies 1: r = `r exp1_descriptives$false_answers_cor_trust_acceptance$estimate`, p `r exp1_descriptives$false_answers_cor_trust_acceptance$p.value`; 2: r = `r exp2_descriptives$false_answers_cor_trust_acceptance$estimate`, p `r exp2_descriptives$false_answers_cor_trust_acceptance$p.value`; 3: r = `r exp3_descriptives$false_answers_cor_trust_acceptance$estimate`, p `r exp3_descriptives$false_answers_cor_trust_acceptance$p.value`; 4: r = `r exp4_descriptives$false_answers_cor_trust_acceptance$estimate`, p `r exp4_descriptives$false_answers_cor_trust_acceptance$p.value`) and only in Study 2 evidence that it is associated with (less) conspiracy beliefs and (less) conspiracy thinking (Studies 1: r = `r exp1_descriptives$false_answers_cor_conspiracy_acceptance$estimate`, p `r exp1_descriptives$false_answers_cor_conspiracy_acceptance$p.value`; 2: r = `r exp2_descriptives$false_answers_cor_conspiracy_acceptance$estimate`, p `r exp2_descriptives$false_answers_cor_conspiracy_acceptance$p.value`; 3: r = `r exp3_descriptives$false_answers_cor_conspiracy_acceptance$estimate`, p `r exp3_descriptives$false_answers_cor_conspiracy_acceptance$p.value`; 4: r = `r exp4_descriptives$false_answers_cor_conspiracy_acceptance$estimate`, p `r exp4_descriptives$false_answers_cor_conspiracy_acceptance$p.value`).

@@

After providing their consent to participate in the study, participants were given an attention check "While watching the television, have you ever had a fatal heart attack?" [1-6; 1 = Never, 6 = Often]. All participants who did not answer "1 = Never" were excluded. Participants then read the following instructions:"We will ask you 10 questions about science. After each question, we will provide you with the scientifically consensual answer and ask whether you accept it." Next, participants answered a set of 10 basic science questions in random order. After each question, participants were presented with an answer reflecting the scientific consensus, and asked whether they accepted it. In Studies 2 and 3, participants additionally saw a short explanation, partly based on explanations generated by ChatGPT, and three links to authoritative sources supporting the answer. In Study 4, we provided only two links and no explanation. Participants then answered questions on conspiracy thinking, conspiracy beliefs, and trust in science.

@@

After removing failed attention checks, we were left with a total sample size of `r exp1_descriptives$n_subj + exp2_descriptives$n_subj + exp3_descriptives$n_subj + exp4_descriptives$n_subj` (`r exp1_descriptives$n_subj` in Study 1, 6 failed attention checks; `r exp2_descriptives$n_subj` in Study 2, 11 failed attention checks; `r exp3_descriptives$n_subj` in Study 3, no failed attention checks; `r exp4_descriptives$n_subj` in Study 4, 2 failed attention checks) participants in the US via prolific. Details and demographics can be found in the online supplemental material. While samples for Studies 1 and 2 were convenience samples, Studies 3 and 4 were conducted on a sample holding vaccine-skeptic beliefs. Prolific Academic allows selecting participants based on their answers to a range of questions. We picked three of these questions and only recruited participants who met our criteria for each of them:

1.  "Please describe your attitudes towards the COVID-19 (Coronavirus) vaccines: [For (I feel positively about the vaccines); Against (I feel negatively about the vaccines); Neutral (I don't have strong opinions either way); Prefer not to say"]. We selected pariticpants who answered "Against".
2.  "Have you received a coronavirus (COVID-19) vaccination? [Yes (at least one dose); No; Prefer not to answer]". We select only people who answered "No".
3.  "On a scale from 1-7, please rate to what extent you agree with the following statement: I believe that scheduled immunizations are safe for children. [1 (totally disagree); 2 (disagree); 3 (somewhat disagree); 4 (neither agree nor disagree); 5 (somewhat agree); 6 (agree); 7 (totally agree); rather not say]". We select only people who answered "1", "2", or "3".

@@

We selected 10 science/health related conspiracy theories from the Belief in Conspiracy Theory Inventory (BCTI) [@pennycookOverconfidentlyConspiratorialConspiracy2022] (Table \@ref(tab:conspiracy)). Participants were asked: "Below is a list of events for which the official version has been disputed. For each event, we would like you to indicate to what extent you believe the cover-up version of events is true or false. [1-9; labels: 1 - completely false, 5 - unsure, 9 - completely true]".

@@

```{r child = "appendix_acceptance_knowledge.Rmd"}
```
